Functional Pearl \\ Programming with Negative and Fractional Types

### 2012-02-19 dual.tex
intuition

\title{Functional Pearl: \\ Programming with Negative and Fractional Types} 
\authorinfo{Roshan P. James}
           {Indiana University}
           {rpjames@indiana.edu}
\authorinfo{Amr Sabry}
           {Indiana University}
           {sabry@indiana.edu}
\maketitle

\begin{abstract}
  Every functional programmer knows about sum and product types, $a+b$ and
  $a*b$ respectively. Negative and fractional types, $a-b$ and $a/b$
  respectively, are much less known and their computational interpretation is
  unfamiliar and often complicated. We show that in a programming model in
  which information is preserved (such as the model introduced in our recent
  paper on \emph{Information Effects}), negative and fractional types have
  particularly intuitive and natural computational
  interpretations. Intuitively, values of negative types are values that flow
  ``backwards'' to satisfy demands, and values of fractional types are values
  that represent first-class pattern clauses. The combination of negative and
  fractional types allow a programmer to express a wide range of programming
  idioms, including higher-order functions, delimited continuations,
  speculative computation, logic programming, and more. 
\end{abstract}

@@

\keywords Arrows, Linear logic, Quantum computing, Reversible logic.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Value Producers and Consumers}

Consider the following ways of purchasing an item that costs \$20.00:

\begin{enumerate}
\item You produce a \$20.00 bill and give it to the seller.
\item You use a credit card. 
\end{enumerate}
In both cases the seller receives the money immediately but there is a subtle
difference. In the second transaction, nobody has yet produced the money that
the seller has received: instead a \emph{debt} is generated and this debt
travels ``backwards'' towards the bank and (hopefully) reconciled there.

Computationally we model the first transaction using essentially the identity
function which receives a \$20.00 bill as input from the buyer and propagates
it on its output to the seller. The second transaction is more
complicated. We model it as shown in the circuit below:

@@

There are two ways to understand this circuit that are both quite
instructive. Let's first examine the type structure of each of the
combinators that comprise the circuit. The first combinator on the right
outputs \$20.00 to the seller. This \$20.00 is produced from nothing so to
speak by generating an equivalent debt that travels backwards. COMPLETE BASED
ON THE FIGURE.

The other way is to follow the execution. It goes forward in time so to speak,
comes back, and then goes forwards again.

\paragraph*{Filinski~\cite{Filinski:1989:DCI:648332.755574}.}
In his Masters thesis, Filinski proposes that continuations are a
\emph{declarative} concept. He, furthermore, introduces a symmetric extension
of the $\lambda$-calculus in which values and continuations are treated as
opposites. This is essentially what we are proposing with one fundamental
difference: our underlying language is not the $\lambda$-calculus but a
language of pure isomorphisms in which information is preserved. This shift
of perspective enables us to distill and generalize the duality of values and
continuations: in particular, in the conventional $\lambda$-calculus setting
values and continuations can be erased and duplicated which makes it
difficult to maintain the correspondence between a value and its negative
counterpart. In contrast, in our setting, one can start from the empty type
$0$, introduce a positive value and its negative counterpart, and let each of
these flow in arbitrary ways. The entire framework guarantees that neither
the value nor its negative counterpart will be deleted or duplicated and
hence that, in any closed program, the ``debt'' corresponding to the negative
value is paid off exactly once.

The forward and backward executions in our framework correspond to
call-by-value and call-by-name. This duality was observed by Filinski and
others following him but it is particularly clean in our framework.

@@

%%%%%%%%%%%%%%%%%
\section{Rules for {{eta}} and {{eps}} }

@@

\section{Diagrams}

@@

>>>>

\title{Functional Pearl \\ Programming with Negative and Fractional Types} 
\authorinfo{Roshan P. James}
           {Indiana University}
           {rpjames@indiana.edu}
\authorinfo{Amr Sabry}
           {Indiana University}
           {sabry@indiana.edu}
\maketitle

\begin{abstract}
  Every functional programmer knows about sum and product types, $a+b$ and
  $a*b$ respectively. Negative and fractional types, $a-b$ and $a/b$
  respectively, are much less known and their computational interpretation is
  unfamiliar and often complicated. We show that in a programming model in
  which information is preserved (such as the model introduced in our recent
  paper on \emph{Information Effects}), negative and fractional types have
  particularly intuitive and natural computational
  interpretations. Intuitively, values of negative types are values that flow
  ``backwards'' to satisfy demands, and values of fractional types are values
  that represent first-class constraints. The combination of negative and
  fractional types allow a programmer to express a wide range of programming
  idioms, including higher-order functions, delimited continuations,
  speculative computation, logic programming, and more.
\end{abstract}

@@

\keywords continuations, information flow, linear logic, logic programming,
quantum computing, reversible logic, symmetric monoidal categories, compact
closed categories.

@@

Connections to type logical grammars and Lambek calculus

Some background on {{langRev}} because it is mentioned in the next section.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Negative Types: Intuition}

Consider the following two ways of purchasing an item that costs \$20.00:
\begin{enumerate}
\item You give the seller a \$20.00 bill.
\item You use a credit card to pay the seller.
\end{enumerate}
In both cases, the seller receives the money immediately but there is a
subtle difference. In the first transaction, the money received by the seller
corresponds to a value that has been produced earlier in the computation. In
the second transaction, the money may or may not exist yet: computationally,
a \emph{debt} is generated to compensate for the money received by the seller
and this debt travels ``backwards'' towards the bank and is (hopefully)
reconciled there.

The example suggests that the ability to consider values flowing backwards
enriches our computational model. This observation goes back to Filinski's
Masters thesis where continuations are identified with these negative values.
We discuss the connections to Filinski's work and others in more detail in
Sec.~\ref{sec:related}. For now, we note that in a conventional programming
language in which values can be copied and deleted at will, extreme care is
needed to keep track of negative values. Indeed, in the example above, if it
were possible to simply delete the variable corresponding to the debt, we
would have produced money out of nothing.

For this reason, our treatment of negative values in the context of
{{langRev}} will be particularly simple. We will have a type $0$ and an
isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
allows us to create a value and a corresponding debt out of nothing. Both the
value and its negative counterpart can flow anywhere in the system. Because
information is preserved, a closed program (which does not have any
``dangling'' negative value) will eventually have to match the negative value
with some corresponding value, effectively using the isomorphism in the
right-to-left direction. In more detail, we can model the credit card
transaction above using the following program (shown diagrammatically):

@@

CLEAN UP the following based on the figure. In contrast, in our setting, one
can start from the empty type $0$, introduce a positive value and its
negative counterpart, and let each of these flow in arbitrary ways. The
entire framework guarantees that neither the value nor its negative
counterpart will be deleted or duplicated and hence that, in any closed
program, the ``debt'' corresponding to the negative value is paid off exactly
once. Computationally we model the first transaction using essentially the
identity function which receives a \$20.00 bill as input from the buyer and
propagates it on its output to the seller. The second transaction is more
complicated. We model it as shown in the circuit below: There are two ways to
understand this circuit that are both quite instructive. Let's first examine
the type structure of each of the combinators that comprise the circuit. The
first combinator on the right outputs \$20.00 to the seller. This \$20.00 is
produced from nothing so to speak by generating an equivalent debt that
travels backwards. COMPLETE BASED ON THE FIGURE. The other way is to follow
the execution. It goes forward in time so to speak, comes back, and then goes
forwards again.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Fractional Types: Intuition}

The type $a/b$ is prominent in the Lambek-Grishin calculus which is
extensively used in computational linguistics. In the common interpretation,
a value of type $a/b$ is a value of some type $c$ such that when put in a
context of type $b$, the result is a value of type $a$. For example, assuming
contexts are represented as functions, a value of type $\texttt{bool} /
(\texttt{int} \rightarrow \texttt{bool})$ is simply a value of type
$\texttt{int}$. Indeed in this case, putting a value of type $\texttt{int}$
in the context $(\texttt{int} \rightarrow \texttt{bool})$ produces a value of
type $\texttt{bool}$. More generally, one can think of a value of type $a/b$
as a value that imposes a constraint on its context of use: it is a value
that requires its context to be of type $b$ and only if that condition is
satisfied, can the value be considered as a value of type $a$. In the
simplest case, the type $1/b$ just constraints the context to be of type $b$.

By allowing the fractional types to be first-class, we can separate the
generation of constraints from their use. For example, consider a variation
on the example in the previous where the parent gives a child the \$20.00
bill but requires that the child only spends the money on buying books. This
can achieved by producing a constraint that can only be resolved if the child
indeed buys a book. Technically, this is achieved by adding a new type
isomorphism {{ 1 <-> a * (1/a) }} which when used in the left-to-right
direction allows the creation of a value and a corresponding constraint. Both
the value and the constraint can flow in arbitrary ways during the
computation but eventually in a closed program with no ``dangling''
constraints, the isomorphism should be used in the right-to-left direction to
resolve the constraints.

NEED to work out the example in detail (or perhaps another example that's
better. Not sure if this one makes sense because we have to create the book
and the context at once.???

@@

%%%%%%%%%%%%%%%%%
\subsection{Rules for {{eta}} and {{eps}} }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Diagrams}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Multiplicative Duality in {{langRev}} }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Related Work} 
\label{sec:related}

Filinski proposes that continuations are a \emph{declarative} concept. He,
furthermore, introduces a symmetric extension of the $\lambda$-calculus in
which values and continuations are treated as opposites. This is essentially
what we are proposing with one fundamental difference: our underlying
language is not the $\lambda$-calculus but a language of pure isomorphisms in
which information is preserved. This shift of perspective enables us to
distill and generalize the duality of values and continuations: in
particular, in the conventional $\lambda$-calculus setting values and
continuations can be erased and duplicated which makes it difficult to
maintain the correspondence between a value and its negative counterpart.

\paragraph*{Filinski~\cite{Filinski:1989:DCI:648332.755574}.}
In his Masters thesis, Filinski proposes that continuations are a
\emph{declarative} concept. He, furthermore, introduces a symmetric extension
of the $\lambda$-calculus in which values and continuations are treated as
opposites. This is essentially what we are proposing with one fundamental
difference: our underlying language is not the $\lambda$-calculus but a
language of pure isomorphisms in which information is preserved. This shift
of perspective enables us to distill and generalize the duality of values and
continuations: in particular, in the conventional $\lambda$-calculus setting
values and continuations can be erased and duplicated which makes it
difficult to maintain the correspondence between a value and its negative
counterpart. In contrast, in our setting, one can start from the empty type
$0$, introduce a positive value and its negative counterpart, and let each of
these flow in arbitrary ways. The entire framework guarantees that neither
the value nor its negative counterpart will be deleted or duplicated and
hence that, in any closed program, the ``debt'' corresponding to the negative
value is paid off exactly once. The forward and backward executions in our
framework correspond to call-by-value and call-by-name. This duality was
observed by Filinski and others following him but it is particularly clean in
our framework.
